---
title: "Wait list from scratch"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simK)
```

Simulating a kidney transplant waiting list is easy to do, with functions from {simK}.

## Candidates

The minimum information needed for kidney transplant candidates encompasses: HLA typing, ABO blood group, cPRA, age, time on dialysis, and clinical urgency.

We can start by defining our wait list size (`n`) and patients' HLA according to its `origin`:

```{r}
candidates <- hla_sample(n = 100, replace = TRUE, origin = 'API')

# lets create also an unique identifier for each patient
candidates$ID <- paste0('R', rownames(candidates))

candidates
```


Patients' blood group can be added with `abo()` and their ages with function `ages()`:

```{r}
# we just have to define sample size and probabilities for A, AB, B and O groups (in this order)
candidates$bg <- abo(n = nrow(candidates), probs = c(0.4, 0.04, 0.06, 0.5))

# for patients' ages we define a trimmed normal distributed value with mean and sd
candidates$age <- ages(n = nrow(candidates), lower = 18, upper = 80, mean = 65, sd = 30)

# let's take a look at our transplant candidates
candidates
```

Now we must define groups of sensitization to transplant candidates according to a 4 elements vector of probabilities corresponding to cPRA = 0%; 1% < cPRA < 50%; 50% < cPRA < 85%; and cPRA > 85%, respectively. 

```{r}
candidates$cPRA <- cpra(n = nrow(candidates), probs = c(0.7, 0.1, 0.1, 0.1))

# and we can also classify those hipersensitised patients
candidates$hiper <- candidates$cPRA > 85
candidates$cPRA <- cpra()
```

Time on dialysis (in months) is computed according to blood group and hipersensitation status:

```{r}
candidates <- candidates |> 
  dplyr::rowwise() |>
  dplyr::mutate(dialysis = dial(hiper = hiper, bg = bg, seed.number = NA)) |>
  dplyr::ungroup()

candidates
```

And we can finish it defining patients clinical urgent (if any):

```{r}
# let´s asume we have 5% of candidates that are clinical urgent for transplantation
candidates$urgent <- sample(c(0,1), size = 100, replace = TRUE, prob = c(0.95,0.05))

# reording the columns
candidates |>
  dplyr::select(ID, bg, A1, A2, B1, B2, DR1, DR2, age, cPRA, hiper, dialysis, urgent)
```

