---
title: "Donors from scratch"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simK)
```

## Donors dataset.

We can build from scratch a data frame with a pool of donors using functions from {simK} and defining it's arguments as we pleased.

The minimum caracteristcs for potential deseaced donors are: blood group, HLA typing and age. So, we start with HLA typing and defining pool size (`n`) and it's `origin`:

```{r}
donors <- hla_sample(n = 100, replace = TRUE, origin = 'API')

# here we create a column with an unique identifier:
donors$ID <- paste0('D', rownames(donors))

donors
```

A column with donors blood group can be added with function `abo()`, we just have to give donors pool size and a vector with ABO genotypic frequencies (`probs`):

```{r}
donors$bg <- abo(n = nrow(donors), probs = c(0.46, 0.03, 0.08, 0.43))

donors
```

Vector on `probs` must be corresponde to A, AB, B and O probabilities (in that order).

And now we just have to add a vector with donors' age with function `ages()`:

```{r}
donors$age <- ages(n = nrow(donors), lower = 18, upper = 80, mean = 55, sd = 25)

# and we can order 
donors |> dplyr::select(ID, bg, A1, A2, B1, B2, DR1, DR2, age)
```

